---
title: "Data Analytics Winter Final Project"
author: 
  - Trifanov Matvei
  - Add you full name here
abstract: "Executive Summary. <!--For the final version: Write an executive summary here. Write it when you are mostly finished with the main report. It should summarize the question, with what data and how you answer it, and what the result is.-->"
format:   
  html: 
    toc: true
    toc-depth: 3
    standalone: true
    embed-resources: true
    code-fold: true
    number-sections: true
---

```{r}
#| label: packages-data
#| message: false
#| echo: false
library(tidymodels)
library(tidyverse)
library(patchwork)
library(ggplot2)
library(dplyr)
library(glue)
library(knitr)
library(corrr)
```

## 1. Introduction
This Homework Project works with dataframes in the package `nycflights13` about flights, specifically a sample of domestic flights that departed from the three major New York City airports in 2013.


The data package `nycflights13` contains five tabular data frames. In total there is information about\
`r nrow(airlines)` **airlines**,\
`r nrow(airports)` **airports**,\
`r nrow(flights)` **flights**,\
`r nrow(planes)` **planes**,\
and `r nrow(weather)` **weather** observations.

The core dataframe `flights` has the following variables:

`year`, `month`, `day`: Date of departure.\
`dep_time`, `arr_time`: Actual departure and arrival times (format HHMM or HMM), local time zone.\
`sched_dep_time`, `sched_arr_time`: Scheduled departure and arrival times (format HHMM or HMM), local time zone.\
`dep_delay`, `arr_delay`: Departure and arrival delays, in minutes. Negative times represent early departures/arrivals.\
`carrier`: Two letter carrier abbreviation. See `airlines` dataframe to get the full name.\
`flight`: Flight number.\
`tailnum`: Plane tail number. See `planes` for additional metadata.\
`origin`, `dest`: Origin and destination. See `airports` for additional metadata.\
`air_time`: Amount of time spent in the air, in minutes.\
`distance`: Distance between airports, in miles.\
`hour`, `minute`: Time of scheduled departure broken into hour and minutes.\
`time_hour`: Scheduled date and hour of the flight as a *POSIXct* date. Along with *origin*, can be used to join flights data to `weather` data.

# 2. Data Load

Load the datasets and libraries.
Provide a brief explanation of each dataset.
Code for loading and viewing the data.

```{r}
#| label: data_load
airlines = read.csv("Data/airlines.csv", sep = ";", header = TRUE)
airports = read.csv("Data/airports.csv", sep = ";", header = TRUE)
flights = read.csv("Data/flights.csv", sep = ";", header = TRUE)
planes = read.csv("Data/planes.csv", sep = ";", header = TRUE)
weather = read.csv("Data/weather.csv", sep = ";", header = TRUE)

# structure of flights dataset
str(flights)

# data summary
summary(planes)
```

## 3. Data Cleaning

Inspect for missing values, duplicates, and inconsistencies.
Clean and handle missing data (impute or remove).
Deal with duplicates and ensure the data is tidy.
Describe each step and your decisions.

```{r}
#| label: data_cleaning
# there we check for missing values in all datasets
# We use colSums(is.na()) to count the missing values for each column
cat("Missing values in airlines:\n")
colSums(is.na(airlines))

cat("Missing values in airports:\n")
colSums(is.na(airports))

cat("Missing values in flights:\n")
colSums(is.na(flights))

cat("Missing values in planes:\n")
colSums(is.na(planes))

cat("Missing values in weather:\n")
colSums(is.na(weather))

# Check for duplicate rows in each dataset
# duplicated() returns TRUE for duplicate rows
cat("\nDuplicate rows in airlines: ", sum(duplicated(airlines)), "\n")
cat("Duplicate rows in airports: ", sum(duplicated(airports)), "\n")
cat("Duplicate rows in flights: ", sum(duplicated(flights)), "\n")
cat("Duplicate rows in planes: ", sum(duplicated(planes)), "\n")
cat("Duplicate rows in weather: ", sum(duplicated(weather)), "\n")


# Remove duplicate rows from datasets if any
# unique() removes duplicate rows
airlines_clean = unique(airlines)
airports_clean = unique(airports)
flights_clean = unique(flights)
planes_clean = unique(planes)
weather_clean = unique(weather)

# Remove rows with too many missing values (e.g., >50% missing)
# This ensures we don't include incomplete data
flights_clean = flights_clean[rowSums(is.na(flights_clean)) <= (0.5 * ncol(flights_clean)), ]

# Impute or remove missing values
# Here, we'll remove rows with missing values for simplicity
airlines_clean = na.omit(airlines_clean)
airports_clean = na.omit(airports_clean)
planes_clean = na.omit(planes_clean)

# For weather and flights, we might need some data imputation
# Impute numerical columns with mean
weather_clean$temperature = ifelse(is.na(weather_clean$temperature), 
                                   mean(weather_clean$temperature, na.rm = TRUE), 
                                   weather_clean$temperature)

# Impute flights' missing values in numerical columns with median
# Example: "air_time" (flight time) column
if ("air_time" %in% colnames(flights_clean)) {
  flights_clean$air_time = ifelse(is.na(flights_clean$air_time), 
                                  median(flights_clean$air_time, na.rm = TRUE), 
                                  flights_clean$air_time)
}

# Verify the datasets are cleaned
cat("\nFinal missing values check for flights:\n")
colSums(is.na(flights_clean))

# Clean column names to avoid issues later
# This removes spaces and special characters in column names
colnames(flights_clean) = gsub("[^[:alnum:]_]", "_", colnames(flights_clean))

# Preview cleaned datasets
head(flights_clean)
head(weather_clean)




```

## 4. Exploratory Data Analysis 

Visualize and summarize each dataset.
Identify patterns, relationships, and trends.
Use visualizations (e.g., histograms, scatterplots, bar charts).
Describe insights from the data exploration. 

```{r}
#| label: EDA


```

## 5. Data Preprocessing

Encode categorical variables (if applicable).
Scale/normalize numerical features.
Merge datasets if needed (e.g., join flights with weather data).
Feature engineering (derive new variables if useful).

```{r}
#| label: preprocessing

```

## 6. Supervised Learning


6.1 Regression Task

Choose a continuous target variable (e.g., flight delay time).
Apply regression models (e.g., linear regression).
Evaluate performance (RMSE, R², etc.).
6.2 Classification Task

Select or derive a categorical target variable (e.g., "on-time" vs "delayed").
Apply classification models (e.g., logistic regression, decision trees).
Evaluate performance (accuracy, precision, recall, etc.).

```{r}
#| label: modeling



```

## Insights and Discussion

Compare model performance.
Highlight key findings and insights.
Discuss any limitations in the analysis.



